Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2012-12-02 3.tex
First complete draft!

There are many precedents for the idea that statistical criteria could be used to differentiate \emph{accidental phonotactic gaps}, those which could arise without any antecedent cause, from \emph{structural phonotactic gaps} \citep[e.g.,][]{Fischer-Jorgensen1952,Saporta1955,Saporta1958,Vogt1954}.

``directly determine the mental representation of the phonotactic constraints'' \citep[180]{Frisch2004}.

In a study of consonant co-occurrence restrictions in Gitksan, \citet{Brown2010} claims that \emph{any} statistically significant pattern in the lexicon is one that is internalized by speakers:

@@

Much of the latter-day literature on phonotactic knowledge, while adopting the representational machinery of generative phonology, takes less interest in the question of what generalizations speakers internalize. In a seminal paper, \citet{McCarthy1988} uses a simple statistical technique to validate a generalization about Arabic root consonant co-occurrence facts originally due to \citet{Greenberg1950}. The assumption appears to be that static co-occurrence restrictions which have a statistically significant trace in the lexicon will also be reflected in psycholinguistic tasks such as wordlikeness judgements. While the Arabic restriction discussed by \citeauthor{McCarthy1988} was ultimately investigated experimentally \citep{Frisch2004}, this is the case for only a small minority of the studies inspired by \citeauthor{McCarthy1988}'s study; in most cases, phonotactic generalizations are inferred directly from lexical statistics \citep[e.g.,][]{Anttila2008a,Berkley1994b,Berkley1994a,Berkley2000,Brown2010,Buckley1997,Coetzee2008a,Elmedlaoui1995,GraffInPress,MacEachern1999,Kinney2005,Kawahara2006,Martin2007,Martin2011,Mester1988,Miller-Ockhuizen2003,Padgett1992,Padgett1995,Pozdniakov2007,Yip1989}.

There are several arguments against this inference. First, negative results in formal learnability theory (see \citealt{Yang2012} for a recent review) show that there are whole classes of rules and constraints which cannot be learned under standard assumptions about acquisition (e.g., that the data is finite, can appear in any order, and that speakers do not have to request grammatical judgements from adults to achieve adult-like competence).

Secondly, there are a number of results which statistically reliable patterns are not learned by speakers \citep[e.g.,][]{Becker2011,Hayes2006,Hayes2009,HayesInPress}. This argues against a direct link between statistical criteria and the synchronic reality of phonotactic generalizations. The danger is that grammatical models which closely mimic lexical statistics but which lack psycholinguistic support (e.g., Gitksan: \citealt{Brown2010}; Muna: \citealt{Anttila2008a}, \citealt{Coetzee2008a}; Shona: \citealt[][385]{Hayes2008a}) may overstate the phonotactic generalizations internalized by speakers.

Another type of non-learning occurs when generalizations are statistically reliable but demonstrably incorrect. \citet{Legate2012}, for instance, note that the overwhelming majority of English words have initial primary stress. While this suggests a quantity-insensitive stress system, English stress is in fact quantity-sensitive (see \citealt{Halle1998c} and references therein).

Finally, there are many cases of statistical generalizations with a known diachronic cause but lacking any synchronic reality. One of the earliest arguments of this type was put forth by Saussure (\citeyear[202f.]{CLG}) in a discussion of rhotacism, a diachronic process which changed Old Latin intervocalic \emph{s} to \emph{r}. As a result, Classical Latin intervocalic \emph{s} is rare. However, Saussure denies that this generalization has any status in Classical Latin.\footnote{See \citet{GormanInPressc} and citations therein for a full explication of Saussure's position on rhotacism, and a restatement of the phonological argument in generative terms.}

Similarly, [ʃ] in Modern English derives from Old English [sk] (e.g., \emph{fisc} `fish'), and since long vowels before complex codas are not found in Old English, [ʃ] is rarely preceded by long vowels in word-final syllables to this day according to \citet{Iverson2005}. The type frequencies of short and long vowels before final [ʃ] and a similar segment, [s] are shown in Table \ref{ssh}. As can be seen, long vowels before final [s] are twice as common before final [ʃ]; according to the appropriate statistical test (the Fisher exact test, described below), this is significant. Yet \citet{Iverson2005} label this constraint ``phonologically accidental'' since a millennium of speakers have produced coinages (e.g., \emph{posh}, \emph{swoosh}) and  borrowings (e.g., \emph{douche}, \emph{capiche}) that disregard this generalization.\footnote{In the loanword adaptation literature, there are many reports that statistically reliable generalizations over the native lexicon are ignored in nativization processes \citep[e.g.,][]{Ito1995a,Ito1995b,Ussishkin2003}. On the other hand, \citet{Frisch2001} suggest that novel borrowings in Egyptian Arabic rarely violate co-occurrence restrictions in that language; however, the generalization in question was exceptionless from Proto-Semitic \citep{Ehret1989} through Classical Arabic \citep{Greenberg1950}, so any violations of this generalization are unexpected.}

\begin{table}[ht]
\centering
\begin{tabular}{l r r r r}
\toprule
          & \{ɪ, ɛ, æ, ʌ, ʊ\}\gap{}\# & \{i, e, ɑ, ɔ, u\}\gap{}\# & \% long & $p$-value \\
\midrule
\gap{}ʃ\# & 78                & 9                 & 8\%      & \multirow{2}{*}{.026} \\
\gap{}s\# & 410               & 107               & 16\%     & \\
\bottomrule
\end{tabular}
\caption{Type frequencies of short and long vowels before word-final [ʃ] and [s] in a subset of the CMU pronunciation dictionary containing all words with token frequency greater than 1 per million words in the SUBTLEX-US frequency norms; the 9 words containing ending in a long vowel followed by ʃ are \emph{douche}, \emph{leash}, \emph{gosh}, \emph{josh}, \emph{posh}, \emph{squash}, \emph{unleash}, \emph{wash}, and \emph{woosh}.}
\label{ssh}
\end{table}

This chapter is a comprehensive investigation of three phonotactic generalizations in Turkish, comparing lexical statistics and the results of a wordlikeness task performed by \citet{Zimmer1969}. Both the lexical statistics and \citeauthor{Zimmer1969}'s experimental results merit reconsideration, because prior discussions do not relate the lexical statistics to competing formalizations of the generalizations involved, and there is widespread disagreement concerning the experimental findings.

One of the three generalizations, \textsc{Labial Attraction}, is shown to   state a statistically reliable generalization about the Turkish lexicon, but has no effect on speakers' wordlikeness judgements. This further undermines the widespread use of statistical criteria to determine which of the vast number of statistically reliable phonotactic generalizations are internalized by speakers and which are not. The simplest hypothesis is that \textsc{Labial Attraction} is inactive in wordlikeness tasks because it is static, corresponding to no phonological process in the language.

@@

Secondly, rather than the use of an unbounded number of Greek-letter variables ($\alpha$, $\beta$, etc.) over feature values $\{+, -\}$, only a single variable, denoted by `$=$', is used \citep{McCawley1973}. A structural description [$=$F]\ldots{}[$=$F] matches a string S$_i$\ldots{}S$_j$ if and only if S$_i$ and S$_j$ are both [$+$F] or both [$-$F]. This is more restrictive than Greek-letter notation, in that it prevents the value of one feature being applied to another. \citet{Odden2012} argues that the two counterexamples against this restriction adduced in \emph{SPE} (352-353) are not probative.

\subsection{Backness harmony}

\citeauthor{Lees1966b} (\citeyear[35]{Lees1966b}, \citeyear[284]{Lees1966a}) models the Turkish vowel harmony system with three rules. The most general of these rules spreads the specification [\textsc{Back}] rightward.

@@

A few complications arise, however. First, as shown in (\ref{turkexcept}a), not all polysyllabic roots conform to \textsc{Backness Harmony}. In this case, suffix vowels generally exhibit harmony with the final root vowel. There is also a very small class of nouns, shown in (\ref{turkexcept}b), which take \emph{-ler}, although their final root vowel is [$+$\textsc{Back}]. Interestingly, they may themselves be harmonic.

@@

\citet[212]{A74} and \citet{Iverson1978} argue that suffix harmony in disharmonic roots found in (\ref{turkexcept}a) requires the rule governing suffix harmony alternations to be distinguished from a sequence structure constraint governing root harmony. Since the rule and sequence structure constraint are otherwise identical, this constitutes an undesirable ``duplication'' in the sense of \citealt{Kisseberth1970b}. However, the theory of exceptionality proposed in \emph{SPE} can account for these facts without duplication \citep[197f.]{Zonneveld1978}.\footnote{\citet[29f.]{Kiparsky1968} discusses a parallel case in Finnish, and proposes a similar separation between the exception-filled sequence structure constraint and an exceptionless suffix harmony process. \citet[171f.]{Howard1972} correctly observes that this duplication is unnecessary.}

\citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention. A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. In other words, no form is never truly an ``exception''; rather, some underlying representations have non-default features which do not match the extended structural description of $R$ which requires that the target be [$+R$]. If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

@@

\noindent
Under the assumptions so far, there is no reason to think that \emph{kabahat} is [$+$\textsc{Backness Harmony}], however: even roots with all back vowels need not undergo this rule. Whatever the ultimate analysis of those few words which fail to show suffix harmony, nothing depends on whether or not they are harmonic.

%\begin{example}[Turkish progressives]
%\begin{tabular}{l l l l l}
%   & \emph{1sg. present}  & \emph{1sg. progressive} & \\
%a. & gelirim              & geliyorum               & `come'  \\
%b. & görürüm              & görüyorum               & `see'   \\
%c. & atarım               & atıyorum                & `throw' \\
%d. & bulurum              & buluyorum               & `find'  \\
%\end{tabular}
%\end{example}

It is necessary to dispense with an alternative analysis proposed by \citet{Clements1982} and \citet{Inkelas1997}. This analysis has desirable properties, but a subtlety of it reintroduces the duplication of rule and sequence structure constraint. Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs. \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), whereas backness of vowels in non-initial syllables is predictable in harmonic roots; there are no prefixes in Turkish. \citeauthor{Clements1982} propose that these vowels, as well as harmonizing suffix vowels, are underspecified for backness, whereas the non-initial vowels of disharmonic roots and of certain exceptional suffixes are fully specified. This is schematized below.

@@

One crucial detail is missing from this analysis: \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, perhaps by structure preservation condition \citep{Kiparsky1985}. However, any condition which prevents \textsc{Backness Harmony} from overwriting underlying backness specifications will reintroduce the duplication of sequence structure and phonological generalizations; under this analysis, disharmonic roots are no longer exceptional, despite considerable evidence (reviewed below) that they are marked in Turkish.\footnote{On the other hand, it is possible to interpret the presence of a single backness specification per root as a sort of default. A precedent for this is the surface-oriented interpretation of the tonal Obligatory Contour Principle proposed by \citet[134]{Goldsmith1976} and \citet{Odden1986}, under which adjacent identical tones are automatically attributed to a single underlying tone. However, this is merely a notational variant of the rule exceptionality account in which [$+$\textsc{Backness Harmony}] is the default.} The the underspecification analysis is rejected here on these grounds.

While harmony in non-derived environments can be inferred from the aforementioned suffix alternations, no evidence has yet been presented to show that Turkish speakers internalize the tendency for roots to conform to backness harmony. If Turkish speakers do not attend to this generalization, there is no need for the grammar to account for it. Several other ``external'' facts suggest that this is not the case. The discussion here is not intended to imply uncritical acceptance of evidence from loanword adaptation, language games or psycholinguistic tasks as evidence for phonological grammar, but rather to illustrate additional evidence that is pertinent if the linking hypothesis is correct.\footnote{Thanks to Bert Vaux and Kie Zuraw for bringing these studies to my attention.}

The production of non-native word-initial onset clusters, discussed by \citet{Clements1982} and \citet{Kaun1999}, suggests that loanword adaptation respects \textsc{Backness Harmony}. Some speakers are said to be capable of pronouncing these non-native clusters, but in fast speech the cluster is split by anaptyxis. In most cases, this vowel matches the following root vowel for backness.

@@

Similar evidence comes from a language game discussed by \citet{Harrison2001}. The game is native to the related language Tuvan, where it is used to convey a sense of ``vagueness or jocularity''; it is not indigenous to Turkish, but can be taught quickly to children or adults. In this game, the base is reduplicated and the first vowel of the reduplicant replaced with a [$+$\textsc{Back}] vowel. In (\ref{redupgame}a), the second [$-$\textsc{Back}] vowel of the base is, in the reduplicant, ``reharmonized'' with the inserted [$+$\textsc{Back}] vowel. The disharmonic roots of (\ref{redupgame}b) do not reharmonize.\footnote{A similar contrast between harmonic and disharmonic roots is found in Tuvan \citep{Harrison2001} and in an unrelated Finnish language game \citep{Campbell1986}.}

@@

\noindent 
In the full specification analysis adopted here, reharmonization is the result of \textsc{Backness Harmony} applying within the reduplicant. On the other hand, the lack of reharmonization in the reduplicants of disharmonic roots suggests that the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication.\footnote{There is reason to believe that reduplicants bear lexical diacritics. Kinande reduplication, documented by \citet{Mutaka1990} and further discussed by \citet{Downing2000}, provides such a case. Verb reduplication requires a bisyllabic reduplicant. This clearly has synchronic force in the gramar, since reduplicated monosyllabic roots actually have three copies of the root, and since reduplicated forms of many trisyllable verbs are ineffable. However, a few trisyllabic verbs exceptionally show full reduplication. This is a lexical property, but one that is only seen on the reduplicant.} \citet{Silverman2000} notes that static phonotactic generalizations are not preserved by operations like reduplication and truncation: under this definition, root harmony is not static.

A number of studies have investigated the role of harmony in word-spotting tasks, though to mimic auditory word recognition and segmentation in natural settings. Many of these studies have been carried out in Finnish, which has a vowel harmony system similar to that of Turkish. \citet{Suomi1997} and \citet{Vroomen1998} task Finnish speakers with identifying harmonic disyllables in an auditory stream. When the syllable preceding the disyllabic target has a different backness specification than the target, recognition of the target is facilitated. Presumably, disharmony facilitates the recognition of word boundaries. \citet{Kabak2010} find that Turkish \textsc{Backness Harmony} has a similar effect: Turkish speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by a disharmonic juncture (e.g., \emph{gölü-PAVO}) than when preceded by a harmonic juncture (e.g., \emph{golu-PAVO}). \citet{Kabak2010} find that this effect does not obtain for speakers of French, a language which lacks vowel harmony. This implies that Turkish speakers have internalized the tendency of harmonic sequences to be root-internal and of disharmonic transitions to cross word boundaries.

The Turkish word-spotting experiment is adapted for infants by \citet{Kampen2008}. In this study, 9-month-old infants are familiarized with an auditory stream containing harmonic disyllabic nonce words. At test time, the infants listen to the disyllabic nonce words in isolation using the head turn preference paradigm. Infants acquiring Turkish listen longer to nonce words preceded by a disharmonic juncture during familiarization (e.g., \emph{lo-NETIS}), whereas infants acquiring German, a language which lacks vowel harmony, do not exhibit this preference. Similarly, \citeauthor{Kampen2008} report that Turkish 6-month-old infants prefer to listen to harmonic nonce words such as \emph{paroz} over disharmonic nonce words like \emph{nelok}, but German 6-month-old infants show no such preference.

It is not clear, however, that the segmentation effects are evidence for grammatical generalizations; for instance, while the effect is defined in terms of harmony, it operates over a domain much larger than the prosodic word over which harmony is computed.

@@

This rule is quite similar to \textsc{Backness Harmony}, but imposes an additional restriction, that targets be [$+$\textsc{High}].

@@

\citet{Lees1966b} describes \textsc{Labial Attraction} as a phonological process by which ``a high, short harmonic vowel is rounded in the second syllable of a disyllabic word whose first vowel is /a/, and whose medial consonant cluster contains a labial /p, b, m, v/, and then it is de-harmonified'' (36).

@@

>>>>

It has long been suggested that statistical criteria could distinguish between \emph{accidental phonotactic gaps}, those which could arise without any antecedent cause, and those gaps which are \emph{structural} in nature \citep[e.g.,][]{Fischer-Jorgensen1952,Saporta1955,Saporta1958,Vogt1954}.
Following seminal work by \citet{Mester1988} and \citet{McCarthy1988}, it is now commonly assumed that the phonotactic patterns in lexical entries ``directly determine the mental representation of the phonotactic constraints'' \citep[180]{Frisch2004} acquired by speakers, and therefore phonotactic constraints, whether categorical or gradient, can be directly inferred from statistical analysis of the lexicon.
There is by now an enormous amount of research that adopts this assumption.
Semitic languages, especially Arabic \citep{Coetzee2008a,Frisch2004,McCarthy1988,Pierrehumbert1993} but also Berber \citep{Elmedlaoui1995} and Tigrinya \citep{Buckley1997}, and Austronesian languages, including Javanese \citep{GraffInPress,Mester1988}, Muna \citep{Anttila2008a,Coetzee2008a}, and Samoan \citep{AldereteInPress}, have been of particular interest to linguists adopting this hypothesis; Aymara \citep{GraffInPress}, Cantonese \citep{Yip1989}, English \citep{Berkley1994a,Berkley1994b,Berkley2000,Coetzee2008b,Martin2007,Martin2011}, Gitksan \citep{Brown2010}, Japanese \citep{Kawahara2006}, Ju|'hoansi, \citep{Kinney2005}, Navajo \citep{Martin2007,Martin2011}, Ofo \citep[38f.]{MacEachern1999}, Russian \citep{Padgett1991,Padgett1992}, Shona \citep{Hayes2008a}, and Wargamay \citep{Hayes2008a} have also received a statistical treatment. 
\citet{Pozdniakov2007} analyze lexical tendencies in a diverse sample of 30 languages.

\citet{Brown2010} presents a strong form of this hypothesis, implying that \emph{any} statistically significant pattern in the lexicon is one that is internalized by speakers:

@@

It would be a result of great interest were it shown that statistical significance is both necessary and sufficient to identify linguistic generalizations which are internalized by speakers, but there is no reason to grant this assumption with respect to phonotactic knowledge.
Nothing demands that the antecedent cause of a statistically reliable lexical tendency be grammatical. 
There are reasons to suspect that many static phonotactic constraints identified in this manner have no synchronic reality.

By hypothesis, the phonological component may impose constraints on sound segments and sequence on underlying representations.
Beyond this, 
a principled null hypothesis (implied by the principle of \textsc{No Static Phonotactics}) is that there is no synchronic, grammatical explanation for underlying representations a language chooses to instantiate.
It is a practical necessity that numerous well-formed underlying representations will be uninstantiated: the lexicon is finite but there are an infinitude of possible URs.
The synchronic grammar cannot reasonably be expected to account for all non-existent underlying representations: for instance, the phonology of English has little to say about absence of */blɪk/ (cf.~\emph{flick}, \emph{brick}, \emph{block}, \emph{blink}).

This much seems uncontroversial.
However, what has not been appreciated is that even when a phonotactic generalization over underlying representations can be given a phonological characterization, there is a plausible alternative to the assumption that it is part of the synchronic grammar: namely, the generalization may be the result of now-complete diachronic change.
Since sound changes often begin as phonological processes, it is no surprise that phonotactic gaps or tendencies can be explained with reference to phonological representations.
But this suggests that the structural nature of a gap is not pertinent to determining whether the constraint is synchronically real.

Saussure (\citeyear{CLG}) presents an example of phonotactic underrepresentation caused by sound change.
With only sporadic exceptions, Old Latin intervocalic \emph{s} undergoes a conditioned phonemic merger with \emph{r}.
This has two consequences.
Second, it introduces many \emph{s}-\emph{r} alternations: e.g., \emph{honōs}-\emph{honōris} `honor'.
The traditional analysis (e.g., \citealt[62]{Foley1965}, \citealt[142]{Gruber2006}, \citealt[134]{Heslin1987}, \citealt[377]{Kenstowicz1996}, \citealt[314]{Klausenburger1976}, \citealt[19]{Matthews1972a}, \citealt[88]{Roberts2012}, \citealt[526]{Watkins1970}) treats \emph{r} as the intervocalic allophone of /s/, and derives \emph{honōris} from underlying /honoːs-is/.

\begin{example}[\textsc{Rhotacism}]
$\textrm{s}~\goesto~\textrm{r}~/~\begin{bmatrix} +\textsc{Voc} \end{bmatrix}~\gap~\begin{bmatrix} +\textsc{Voc} \end{bmatrix}$
\end{example}

\noindent
However, subsequent sound changes \citep[e.g.,][]{Baldi1994,Safarewicz1932}, particularly the degemination of Old Latin \emph{ss} before diphthongs and long monophthongs (e.g., \emph{caussa} $>$ \emph{causa} `cause'), introduce numerous exceptions to \textsc{Rhotacism}.
In Classical Latin, intervocalic \emph{s} is found root-internally (\emph{asellus} `donkey', \emph{casa} `hut'), in environments derived by inflectional suffixes (\emph{uās}-\emph{uāsis} `vase', \emph{uisēre} `to view'), prefixation (\emph{dēsecāre} `to cut off'; cf.~\emph{dē} `from', \emph{secāre} `to cut'), compounding (\emph{olusātrum} `parsnip'; cf.~\emph{olus} `vegetable', \emph{ātrum} `black'), and denominal adjective formation (\emph{uentōsus} `windy'; cf.~\emph{uentus} `wind'), and is tolerated in nativized loanwords from Celtic (\emph{omāsum} `tripe'), Germanic (\emph{glaesum} `amber'), and Greek (\emph{basis} `pedestal').
These facts lead Saussure to conclude that \textsc{Rhotacism} is no longer ``inhérente à la nature de la langue'' (202).
While some of the apparent exceptions may be the result of opaque phonological application \citep{Heslin1987}---an explanation not yet available in Saussure's time---any formulation of rhotacization will admit nearly as many lexical exceptions as there are roots exhibiting \emph{s}-\emph{r} alternations \citep{GormanInPressc}.
Any synchronic account of the underrepresentation of intervocalic \emph{s} must confront and the unproductive nature (as indicated by the accumulation of exceptions) of the proximate explanation for this tendency.

Other cases show that constraints identified by statistical analysis are inert: that is, there is no evidence (of any of type described in \S\ref{s:espt} above) that they are internalized by speakers.
For instance, most instances of Modern English [ʃ] derive from Old English [sk] (e.g., \emph{fisc} `fish') via sound change.
Since Old English does not permit long vowels before complex codas, compared to similar segments like [s], [ʃ] is still rarely preceded by long vowels in word-final syllables.\footnote{
    This sample is drawn from the CMU pronunciation dictionary, filtered by excluding words with a token frequency of less than 1 per million words in the SUBTLEX-US frequency norms.
    Less restrictive samples give similar results.
    The 9 words ending with a long vowel-[ʃ] sequence are \emph{douche}, \emph{leash}, \emph{gosh}, \emph{josh}, \emph{posh}, \emph{squash}, \emph{unleash}, \emph{wash}, and \emph{woosh}.}
As can be seen, long vowels are twice as common before [s] as before [ʃ], a significant generalization according to the Fisher exact test.
Similarly, \citet{HayesInPress} report that this constraint is discovered by the \citet{Hayes2008a} phonotactic learning model, which uses a related statistical criterion to identify constraints.
Despite this, \citet{Iverson2005} label the constraint on long vowels before [ʃ] as ``phonologically accidental'', as a millennium of coinages (e.g., \emph{posh}) and loanwords (e.g., \emph{douche}) disregard this generalization.
\citet{HayesInPress} find that a variant of this restriction has little to no effect on wordlikeness judgements.

\begin{table}[t]
\centering
\begin{tabular}{l r r r r}
\toprule
          & \{ɪ, ɛ, æ, ʌ, ʊ\}\gap{}\# & \{i, e, ɑ, ɔ, u\}\gap{}\# & \% long & $p$-value \\
\midrule
\gap{}ʃ\# & 78                & 9                 & 8\%      & \multirow{2}{*}{.026} \\
\gap{}s\# & 410               & 107               & 16\%     & \\
\bottomrule
\end{tabular}
\caption{Type frequencies of short and long vowels before word-final [ʃ] and [s]}
\label{ssh}
\end{table}

This latter example makes it clear that a purely statistical criterion overgenerates in the sense that it predicts phonotactic constraints which speakers do not seem to internalize.
To account for the cases like the one above, \citet{HayesInPress} propose that speakers are biased in favor of ``natural generalizations'' in probabilistic phonotactic learning.\footnote{
    \citeauthor{HayesInPress} do not provide an operational definition of ``naturalness'', so it is difficult to evaluate their specific results or to determine whether they apply the classification of ``natural'' and ''unnatural'' is applied in an ad hoc (or post hoc) fashion.}
However, this chapter argues that this overgeneration gives the lie to the broader assumption that phonotactics are extracted from patterns in the lexicon.
Rather, the only restrictions which speakers internalize are those which derive from phonological processes in the language.
This is independent of ``naturalness'', since both ``natural'' and ``unnatural'' variants of statistically reliable static constraints are shown to be inert.

This chapter focuses on three phonotactic generalizations in Turkish, comparing lexical statistics and the results of a wordlikeness task performed by \citet{Zimmer1969}. 
Both the lexical statistics and \citeauthor{Zimmer1969}'s experimental results merit reconsideration, because prior discussions do not relate the lexical statistics to experimental data or competing formalizations of the generalizations involved.

@@

Secondly, rather than the use of an unbounded number of Greek-letter variables ($\alpha$, $\beta$, etc.) over feature values $\{+, -\}$, only a single variable, denoted by `$=$', is used \citep{McCawley1973}. A structural description [$=$F]\ldots{}[$=$F] matches a string S$_i$\ldots{}S$_j$ if and only if S$_i$ and S$_j$ are both [$+$F] or both [$-$F]. This is more restrictive than Greek-letter notation, in that it prevents the value of one feature being applied to different feature; \citet{Odden2012} argues that the two counterexamples against this restriction presented in \emph{SPE} (352-353) are not probative.

\subsection{Backness harmony}
\label{ss:bh}

\citeauthor{Lees1966b} (\citeyear[35]{Lees1966b}, \citeyear[284]{Lees1966a}) models the Turkish vowel harmony system with three rules; the most general of these rules spreads the specification [\textsc{Back}] rightward.

@@

A few complications arise, however. First, as shown in (\ref{turkexcept}a), not all polysyllabic roots conform to \textsc{Backness Harmony}. In this case, suffix vowels generally exhibit harmony with the final root vowel. There is also a very small class of nouns, shown in (\ref{turkexcept}b), which take \emph{-ler}, although their final root vowel is [$+$\textsc{Back}]. Interestingly, the roots themselves may be harmonic.

@@

\citet[212]{A74} and \citet{Iverson1978} argue that suffix harmony in disharmonic roots found in (\ref{turkexcept}a) requires the rule governing suffix harmony alternations to be distinguished from a sequence structure constraint governing root harmony. 
Since the rule and sequence structure constraint are otherwise identical, this constitutes a ``duplication''. 
However, the theory of exceptionality proposed in \emph{SPE} can account for these facts without duplication \citep[197f.]{Zonneveld1978}.\footnote{
    \citet[29f.]{Kiparsky1968} discusses a parallel case in Finnish, and proposes a similar separation between the exception-filled sequence structure constraint and an exceptionless suffix harmony process. 
    \citet[171f.]{Howard1972} correctly observes that this duplication is also unnecessary.}

\citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention. 
A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. 
In other words, no representation is never truly an exception; rather, some underlying representations have non-default features which do not match the extended structural description of $R$, which requires that the target be [$+R$]. 
If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

@@

\noindent
Under the assumptions so far, there is no reason to think that \emph{kabahat} is [$+$\textsc{Backness Harmony}], however: even roots with all back vowels need not undergo this rule. Whatever the ultimate analysis of those few words which fail to show suffix harmony, nothing depends on whether or not the roots in question are harmonic.

It is necessary to dispense with an alternative analysis proposed by \citet{Clements1982} and \citet{Inkelas1997}. 
This analysis has desirable properties, but it subtlely reintroduces the duplication of phonological process and sequence structure constraint. 
Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs. \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), whereas backness of vowels in non-initial syllables is predictable in harmonic roots (there are no prefixes in Turkish).
\citeauthor{Clements1982} propose that these vowels, as well as harmonizing suffix vowels, are underspecified for backness, whereas the non-initial vowels of disharmonic roots and of certain exceptional suffixes are fully specified. 
This is schematized below.

@@

One crucial detail is missing from this analysis: \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, perhaps by structure preservation condition \citep{Kiparsky1985}. 
However, any condition which prevents \textsc{Backness Harmony} from overwriting underlying backness specifications will reintroduce the duplication of sequence structure and phonological generalizations; under this analysis, disharmonic roots are no longer exceptional, despite considerable evidence (reviewed below) that they are marked in Turkish.\footnote{
    On the other hand, it is possible to interpret the presence of a single backness specification per root as a sort of default. 
    A precedent for this is the surface-oriented interpretation of the tonal Obligatory Contour Principle proposed by \citet[134]{Goldsmith1976} and \citet{Odden1986}, under which adjacent identical tones are automatically attributed to a single underlying tone. 
    However, this is merely a notational variant of the rule exceptionality account in which [$+$\textsc{Backness Harmony}] is the default.} 
    The underspecification analysis is rejected here on these grounds.

While harmony in non-derived environments can be inferred from the aforementioned suffix alternations, no evidence has yet been presented to show that Turkish speakers internalize the tendency for roots to conform to backness harmony. 
    If Turkish speakers do not attend to this generalization, there is no need for the grammar to account for it. Several other ``external'' facts suggest that this is not the case. 
    The discussion here is not intended to imply uncritical acceptance of evidence from loanword adaptation, language games, or psycholinguistic tasks as evidence for phonological grammar, but rather to illustrate additional evidence that is pertinent if the linking hypothesis is correct.\footnote{
    Thanks to Bert Vaux and Kie Zuraw for bringing these studies to my attention.}

The production of non-native word-initial onset clusters, discussed by \citet{Clements1982} and \citet{Kaun1999}, suggests that loanword adaptation respects \textsc{Backness Harmony}. 
    Some speakers pronounce these non-native clusters, but in fast speech the cluster is split by anaptyxis. 
    In the majority of cases, this vowel matches the following root vowel for backness.

@@

Similar evidence comes from a language game discussed by \citet{Harrison2001}. 
    The game is native to the related language Tuvan, where it is used to convey a sense of ``vagueness or jocularity''; it is not indigenous to Turkish, but can be taught quickly to children or adults. 
    In this game, the base is reduplicated and the first vowel of the reduplicant replaced with a [$+$\textsc{Back}] vowel. 
    In (\ref{redupgame}a), the second [$-$\textsc{Back}] vowel of the base is, in the reduplicant, ``reharmonized'' with the inserted [$+$\textsc{Back}] vowel. 
    The disharmonic roots of (\ref{redupgame}b) do not reharmonize.\footnote{
    A similar contrast between harmonic and disharmonic roots is found in Tuvan \citep{Harrison2001} and in an unrelated Finnish language game \citep{Campbell1986}.}

@@

\noindent 
In the full specification analysis adopted here, reharmonization is the result of \textsc{Backness Harmony} applying within the reduplicant. 
    On the other hand, the lack of reharmonization in the reduplicants of disharmonic roots suggests that the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication.\footnote{
    There is reason to suspect that reduplicants bear the lexical diacritics of the roots they are derived from.
    In Kinande, verb reduplication, 
    Kinande reduplication, documented by \citet{Mutaka1990} and further discussed by \citet{Downing2000}, requires a bisyllabic reduplicant. 
    This clearly has synchronic force in the gramar, since reduplicated monosyllabic roots contain three copies of the root so as to satisfy this requirement, and reduplicated forms of many trisyllabic verbs are ineffable. 
    However, a few trisyllabic verbs exceptionally show full reduplication.
    This is a lexical property, but presence or absense of this exceptionalty feature is only apparent in the form of the reduplicant.}
    Under this definition, root harmony is not static, and so this is not an exception to the claim that static phonotactic generalizations are not preserved by operations like reduplication and truncation \citep{Silverman2000}.

A number of studies have investigated the role of harmony in word-spotting tasks, though to mimic auditory word recognition and segmentation in natural settings. 
    Many of these studies have been carried out in Finnish, which has a vowel harmony system similar to that of Turkish. 
    \citet{Suomi1997} and \citet{Vroomen1998} task Finnish speakers with identifying harmonic disyllables in an auditory stream. 
    When the syllable preceding the disyllabic target has a different backness specification than the target, recognition of the target is facilitated. Presumably, disharmony facilitates the recognition of word boundaries. 
    \citet{Kabak2010} find that Turkish \textsc{Backness Harmony} has a similar effect: Turkish speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by a disharmonic juncture (e.g., \emph{gölü-PAVO}) than when preceded by a harmonic juncture (e.g., \emph{golu-PAVO}). 
    \citeauthor{Kabak2010} report that this effect does not obtain for speakers of French, a language which lacks vowel harmony. 
    This implies that Turkish speakers have internalized the tendency of harmonic sequences to be root-internal and of disharmonic transitions to cross word boundaries.

The Turkish word-spotting experiment is adapted for infants by \citet{Kampen2008}.
In this study, 9-month-old infants are familiarized with recordings of harmonic, disyllabic nonce words presented in isolation.
At test time, the infants listen to the disyllabic nonce words in isolation using the head turn preference paradigm. 
Infants acquiring Turkish listen longer to nonce words preceded by a disharmonic juncture during familiarization (e.g., \emph{lo-NETIS}), whereas infants acquiring German, a language which lacks vowel harmony, do not exhibit this preference. 
Similarly, \citeauthor{Kampen2008} report that Turkish 6-month-old infants prefer to listen to harmonic nonce words such as \emph{paroz} over disharmonic nonce words like \emph{nelok}, but German 6-month-old infants show no such preference.
However, there are some caveats for identifying these effects with grammatical computations: the domain for segmentation effects is considerably larger than the domain for harmony, so the two cannot be easily identified.
All that can be said is that there is an obvious similarity between computing the contexts for vowel harmony and for the word segmentation heuristic.

@@

\textsc{Roundness Harmony} is quite similar to \textsc{Backness Harmony}, but imposes an additional restriction, that targets be [$+$\textsc{High}].

@@

\citet[35]{Lees1966b} notes the tendency of Turkish high back vowels to be round after \emph{a}-labial consonant sequences, and formalizes this as a phonological process.

@@

%\begin{example}[Turkish progressives]
%\begin{tabular}{l l l l l}
%   & \emph{1sg. present}  & \emph{1sg. progressive} & \\
%a. & gelirim              & geliyorum               & `come'  \\
